---
title: "Adelino Pedon Range in Characteristics"
author: "`r Sys.getenv('USERNAME')`"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document: null
params:
  genhz_rules: !r as.character("C:/PedonRIC/adelino.R")
  series: !r as.character("Bluepoint")
---
```{r setup, echo=FALSE, warning=FALSE}
# generic horizon rules
source(params$genhz_rules)

# knit options
knitr::opts_chunk$set(echo=FALSE, results='asis', warning=FALSE, message=FALSE, background="#F7F7F7", fig.retina=1, dev="png", tidy=FALSE, verbose=FALSE, cache = F)

options(stringsAsFactors = FALSE)
```

```{r load packages}
# soil libraries
library(aqp)
library(soilDB)
library(soilReports)

# data manipulation libraries
library(knitr)
library(plyr)
library(reshape2)
library(circular)

# graphic libraries
library(lattice)
library(latticeExtra)
library(RColorBrewer)


# mapping libraries
library(maps)
library(maptools)
library(mapview)

# custom ggplot2like latticeExtra theme
opar <- trellis.par.get()
trellis.par.set(theme = ggplot2like())
tpg <- trellis.par.get()
tpg$axis.line <-  list(alpha = 1, col = "#000000", lty = rep(1, 7), lwd = rep(0.2, 7))
tpg$strip.border <- list(alpha = 1, col = "#000000", lty = rep(1, 7), lwd = rep(0.2, 7))
trellis.par.set(tpg)
```


```{r fetch and format, load-data, echo=FALSE, warning=FALSE}
# load NASIS data
pedons <- fetchNASIS(lab = TRUE)

h <- horizons(pedons)
s <- site(pedons)
d <- diagnostic_hz(pedons)

# modify the horizon table
h <- within(h, {
  hzname = ifelse(is.na(hzname), "missing", hzname)
  hzdepm = (hzdept + hzdepb) / 2
  genhz  = generalize.hz(hzname, ghr$n, ghr$p, hzdepm = hzdepm)
  gravel = gravel - fine_gravel
  })
var <- c("fine_gravel", "gravel", "cobbles", "stones", "boulders", "paragravel", "paracobbles", "channers", "flagstones")
h[var] = lapply(h[var], as.integer)
  
horizons(pedons) <- h


# rename rock fragments
idx <- names(h) == "total_frags_pct"
names(h)[idx] <- c("fragvoltotc")

# vector of names to exclude from numeric summary
vars <- c('peiid', 'phiid', 'hzname', 'genhz', 'hzdept', 'hzdepb', 'labsampnum', 'd_value', 'd_chroma', 'm_value', 'm_chroma', 'parafine_gravel', 'paragravel', 'paracobbles', 'parastones', 'paraboulders', 'parachanners', 'paraflagstones', 'unspecified', 'd_r', 'd_b', 'd_g', 'd_sigma', 'm_r', 'm_b', 'm_g', 'm_sigma')

# create vector of names for numeric soil variables excluding vars vector
num_vars <- names(h)[! names(h) %in%  vars &                     # remove vars
                       sapply(h, is.numeric) &                   # select numeric columns
                       sapply(h, function(x) !all(is.na(x)))  &  # select non-empty columns 
                       sapply(h, function(x) !all(x == 0))       # remove columns were all observations == 0
                     ]

num_vars <- na.omit(num_vars)


# fig.height
nh <- length(num_vars[num_vars %in% names(h)]) / 4 * 3

# modify diagnostic table
# create a empty diagnostic table if data is NULL
if (nrow(d) == 0) {
  d <- data.frame(peiid = s$peiid,
                  diag_kind = as.character("missing"),
                  featdept = as.integer(NA), 
                  featdepb = as.integer(NA)
                  )
  }
d <- transform(d, 
               thickness = featdepb - featdept,
               featkind = as.character(featkind)
               )

# modify site table
# rename psc depths
idx <- names(s) == "psctopdepth" | names(s) == "pscbotdepth"
names(s)[idx] <- c("featdept", "featdepb")

s <- transform(s, thickness = featdepb - featdept)

srf <- s[grepl("surface_", names(s))]
names(srf) <- gsub("surface_", "", names(srf))
srf <- within(srf, {
  total_srf = gravel + cobbles + stones + boulders + flagstones + channers
  gravel = gravel - fgravel
  })
```

# Summary of pedon data

## Interactive Map

```{r map, echo=FALSE, results='asis'}

if (dim(s)[1] != 0) {
  pedon_locations <- s[complete.cases(s[c("x_std", "y_std")]), ]
  coordinates(pedon_locations) <- ~ x_std + y_std
  proj4string(pedon_locations) <- CRS("+init=epsg:4326")
  
  if(params$series != "Generic") {
    series_extent <- seriesExtent(params$series)
    
    mapviewOptions(basemaps = "Esri.WorldImagery")
    
    mapView(pedon_locations) #+ series_extent
    } else mapView(pedon_locations)
  } else("no coordinates")
```

## Site Data

```{r format site data, results='asis', echo=FALSE}
# Site information
kable(subset(s, select = c("pedon_id", "taxonname", "taxsubgrp", "taxpartsize", "pedontype", "describer")), caption = "Summary of data in the selected set")
```


## Plot of Soil Profiles
```{r profile plots, fig.width=8}
par(mar=c(0,0,2,1))
plot(pedons, label='site_id')

```

## Comparison of genhz and hzname pattern matching

```{r genhz matching, fig.width=8}
hz_t <- addmargins(table(h$genhz, h$hzname))

idx <- pIndex(hz_t, 15)

# plot 15 horizon designations per row
for (i in unique(idx)){
  print(kable(hz_t[, c(idx == i)], align = "c", digits = 0, caption = "Horizon designations vs generic horizon designations (counts)"))
  }
```

## Generalized Horizon Plots

```{r, echo=FALSE}
cols <- brewer.pal(n = length(levels(pedons$genhz)), name = "Set1") 
hz.names <- levels(pedons$genhz)
# assign a color to each generalized horizon label
pedons$genhz.soil_color <- cols[match(pedons$genhz, hz.names)]

idx <- pIndex(pedons, 15)

# plot 15 profiles at a time
for (i in unique(idx)){
  plot(pedons[idx == i], name = 'hzname', color = 'genhz.soil_color', label = 'pedon_id')
  title("Soil profile plots")
  #legend('bottomleft', legend = hz.names, pt.bg = cols, pch = 22, horiz = TRUE, pt.cex = 2, text.width = 1)
  }
```

# Range in characteristics (RIC) for NCSS pedon lab data

## Surface rock fragments

```{r surface rock fragments, echo=FALSE, results='asis'}
vars <- c("total_srf", "fgravel", "gravel", "cobbles", "stones", "boulders", "channers", "flagstones")
srf.lo <- melt(srf, measure.vars = vars)
srf.5n <- ddply(srf.lo, .(variable), summarize,
                range = prettySummary(value, p=c(0,0.5,1))
                )

kable(srf.5n, align = "c", caption =  "Surface rock fragments (min, median, max)(number of pedons)")
 
#if (sum(srf$total_srf, na.rm = T) != 0) {
#  bwplot(variable ~ value, data = srf.lo, 
#         main = "Boxplots of surface rock fragments",
#         ylab = "percent",
#         axis = axis.grid
#         )
#  }         
```

## Diagnostic horizons and soil characteristics

```{r diagnostics, echo=FALSE, results='asis', warning=FALSE}
diag.lo <- melt(d, id.vars = "featkind", measure.vars = c("featdept", "featdepb", "thickness"), factorsAsStrings = FALSE)
pscs.lo <- melt(s, id.vars = "peiid", measure.vars = c("featdept", "featdepb", "thickness"), factorsAsStrings = FALSE)
pscs.lo <- data.frame(featkind = "particle size control section", 
                      variable = pscs.lo$variable, 
                      value = pscs.lo$value
                      )

# combine diagnostic and particle size control section long tables
diag.lo <- rbind(diag.lo, pscs.lo)
diag.5n <- ddply(diag.lo, .(variable, featkind), summarize,
                range = prettySummary(value, p=c(0,0.5,1))
                )
diag.wi <- dcast(diag.5n, featkind ~ variable, value.var = 'range')

kable(diag.wi, align = "c", caption = "Depths and thickness of diagnostic horizons and features(min, median, max)(number of pedons)")


#if (!all(is.na(diag.lo$value))) {
#  bwplot(featkind ~ value | variable, data = diag.lo, 
#         main = "Boxplots of diagnostic horizon and feature depths", 
#         scales =list(x="free"), axis = axis.grid, 
#         as.table = TRUE
#         )
#  }
```


## Generalized horizon depths

```{r hzname vs genhz, echo=FALSE, results='asis'}
genhz.lo <- melt(h, id.vars="genhz", measure.vars = c('hzdept', 'hzdepb'))
genhz.thk <- ddply(h, .(phiid, genhz), summarize, thickness=sum(hzdepb-hzdept))
genhz.lo2 <- melt(genhz.thk, id.vars = "genhz", measure.vars = 'thickness')
genhz.lo <- rbind(genhz.lo, genhz.lo2)
genhz.5n <- ddply(genhz.lo, .(variable, genhz), summarize,
                range = prettySummary(value, p=c(0,0.5,1))
                )

kable(dcast(genhz.5n, genhz ~ variable, value.var = 'range'), align = "c", caption = "Depths and thickness of generic horizons (min, median, max)(number of horizons)")

genhz.lo$genhz <- factor(genhz.lo$genhz, levels = rev(levels(genhz.lo$genhz)))

#bwplot(genhz ~ value | variable, data = genhz.lo, 
#       main = "Boxplots of horizon generic horizon depths and thickness",
#       scales =list(x="free"), axis = axis.grid,
#       as.table = TRUE
#       )
```


## Range in characteristics (RIC) for static soil properties 

### Tables

```{r ric ssp tables}
h.lo <- melt(h, id.vars="genhz", measure.vars = num_vars)
h.5n <- ddply(h.lo, .(variable, genhz), summarize,
                range = prettySummary(value, p=c(0,0.5,1))
                )
h.wi <- dcast(h.5n, genhz ~ variable, value.var = 'range')

idx <- pIndex(h.wi, 5)

for (i in unique(idx)) {
  print(kable(h.wi[, c(T, idx == i)], align = "c", caption = "Numeric variables by generic horizon (min, median, max)(number of horizons)"))
  # inserting an empty line so the last table doesn't come out corrupted
  cat("\n")
  }

h.lo$genhz <- factor(h.lo$genhz, levels = rev(levels(h.lo$genhz)))
n <- ceiling(length(levels(h.lo$variable))/4)


### Boxplots
#r ric ssp boxplots, fig.height=nh, fig.width=8}
#bwplot(genhz ~ value | variable, data = h.lo,
#       main = "Box plots of numeric variables by generic horizon",
#       scales=list(x="free"), axis = axis.grid,
#       as.table = TRUE, layout = c(4, n)
#       )
```

## Chemistry
```{r chemistry}
# vector containing which variables are being selected
chem.vars <- c('phfield', 'ec_lab')
# data transformation from wide to long format
h.chem.1 <- melt(h, id.vars = "genhz", measure.vars = chem.vars)
# summarize data values by genetic horizon
h.chem <- ddply(h.chem.1, .(variable, genhz), summarize, range=prettySummary(value, p=c(0,0.5,1)))
# aggregate data back by genetic horizon
h.chem <- dcast(h.chem, genhz ~ variable, value.var = 'range')
kable(h.chem, align = 'c', caption = "Chemical properties by genetic horizon (min, median, max)(n)")
```

## Effervescence
```{r effervescence}
# changing name 
h$effervescence[h$effclass == "very slight"] <- "very.slight"
# storing levels as vector
eff <- c(violent = "violent", strong = "strong", slight = "slight", very.slight = "very slight", none = "none", missing = "missing")
# applying lables as a factor
h$effervescence <- eff[h$effervescence]
effclass <- factor(h$effervescence, levels = eff)
# create table of effervescence class by genetic horizon
effclass.t <- xtabs(~ genhz + effclass, data = h, drop.unused.levels = FALSE)
# viewing table
#effclass.t
# creating table with ordered cols
#kable(addmargins(effclass.t[, c(5,1,2,3,4)]), caption = "Effervescence by genetic horizon counts)", align = 'c')
kable(addmargins(xtabs(~ genhz + effclass, data = h, drop.unused.levels = TRUE)), digits = 0, caption = "Effervescence by generic horizon", align = 'c')
```


## Salinity
```{r Salinity}
# assigning salinity class based on ec data directly to new saltclass slot
h$saltclass <- ifelse(
  h$ec_lab < 2, "nonsaline",
  ifelse(
      h$ec_lab < 4, "very.slightly",
      ifelse(
          h$ec_lab <8, "slightly",
          ifelse(
              h$ec_lab < 16, "moderately",
              ifelse(
                  h$ec_lab >= 16, "strongly", 
                  "na"
                )
            )
        )
    )
)
# vector containing salt classes
salt <- c(nonsaline = "nonsaline", very.slightly = "very slightly", slightly = "slightly", moderately = "moderately", strongly = "strongly", na = "na")
# applying names to salt class 
#h$saltclass <- salt[h$saltclass]
h$saltclass <- factor(h$saltclass, levels =salt)
# create salt class table gy genetic horizon
#saltclass.t <- xtabs(~ genhz + saltclass, data = h, drop.unused.levels = FALSE)
# reorder table columns
#saltclass.t[ , c(2,5,3,1,4)]
# final table with reorderd cols 
#kable(addmargins(saltclass.t[, ]), caption = "Salinity Class by genetic horizon (counts)", align = 'c')

  
if(is.na(h$saltclass)){
  print("ec not populated")} else
  {kable(addmargins(xtabs(~ genhz + saltclass, data = h, drop.unused.levels = TRUE)), digits = 0, caption = "Effervescence by generic horizon", align = 'c')
  }
```


## Alkalinity (pH Class)
```{r alkalinity}
# converting ph values to clas
h$phclass <- ifelse(
    h$phfield < 6.0, "moderately.acid",
    ifelse(
      h$phfield < 6.5, "slightly.acid",
      ifelse(
        h$phfield < 7.3, "neutral",
        ifelse(
          h$phfield < 7.9, "slightly.alkaline",
          ifelse(
            h$phfield < 8.4, "modeately.alkaline",
            ifelse(
              h$phfield <= 9, "strongly.alkaline",
              ifelse(
                h$phfield > 9, "very.strongly.alkaline",
                "na"
              )
            )
          )
        )
      )
    )
)
# creating a vector of levels
ph <- c(moderately.acid = "moderately acid", slightly.acid = "slightly acid", neutral = "neutral", slightly.alkaline = "slightly alkaline", moderately.alkaline = "moderately alkaline", strongly.alkaline = "strongly alkaline", very.strongly.alkaline = "very strongly alkaline", na = "none")
# applying names
#h$phclass <- ph[h$phclass]
#phclass <- factor(h$phclass, levels = ph)

if(is.na(h$phclass)){print("pH not populated")} else
{kable(addmargins(xtabs(~ genhz + phclass, data = h, drop.unused.levels = TRUE)), digits = 0, caption = "pH Class by generic horizon", align = 'c')}
```

### Texture

```{r texture}
kable(addmargins(xtabs(~ genhz + texcl, data = h, drop.unused.levels = TRUE)), digits = 0, caption = "Texture by generic horizon (counts)")

hz_tex <- addmargins(xtabs(~ genhz + texture, data = h))
idx <- pIndex(hz_tex, 15)

for (i in unique(idx)){
  print(kable(hz_tex[, c(idx == i)], align = "c", digits = 0, caption = "Tex Mod & Class by generic horizon (counts)"))
  }
```

### Texture Triangles
```{r text triangle}
# create soil texture tiangle plots
#library(soiltexture)


#texture.df <- cbind.data.frame(pedons@horizons$genhz, pedons@horizons$sand, pedons@horizons$silt, pedons@horizons$clay)
#colnames(texture.df) <- c("genhz", "SAND", "SILT", "CLAY")
#texture.df <- texture.df[complete.cases(texture.df),]

#texture.df$genhz <- droplevels(texture.df$genhz)
#texture.df <- droplevels(texture.df[!texture.df$genhz == 'not-used',])


#gennames <- unique(unlist(texture.df$genhz))
#gennames <- droplevels(gennames)

# loop through for plots
#for(i in 1:length(gennames)){
#  h.text <-subset(texture.df, genhz==gennames[i])
#       TT.plot( class.sys = "USDA-NCSS.TT",
#         tri.data = h.text, 
#         main = c(levels(gennames)[i]," Horizon Soil #Texture"),#c(#gennames, "Soil Texture"), 
#         col = "red", 
#         cex = .5, 
#         grid.show = FALSE,
#         cex.axis = .5,
#         cex.lab = .7)
#}


```

### Color

```{r color}
kable(addmargins(xtabs(~ h$genhz + h$d_hue, data = h, drop.unused.levels = TRUE)), digits = 0, caption = "Dry hue by generic horizon (counts)")

kable(addmargins(xtabs(~ genhz + m_hue, data = h, drop.unused.levels = TRUE)), digits = 0, caption = "Moist hue by generic horizon (counts)")

# Value and Chroma
col.vars <- c('d_value', 'd_chroma', 'm_value', 'm_chroma')
h.col <- melt(h, id.vars = "genhz", measure.vars = col.vars)
h.col <- ddply(h.col, .(variable, genhz), summarize, range=prettySummary(value, p=c(0,0.5,1)))
h.col <- dcast(h.col, genhz ~ variable, value.var = 'range')
kable(h.col, align = 'c', caption = "Color(Value and Chroma) by genetic horizon (min, median, max)(n)" )


```

### Effervescence

```{r effervesence}
kable(addmargins(xtabs(~ genhz + effclass, data = h, drop.unused.levels = TRUE)), digits = 0, caption = "Effervescence by generic horizon (counts)")
```


## Range in characteristics (RIC) for the geographic setting

## Landform
```{r, landscape and land form}
library(Hmisc)
lf <- sort(table(pedons$landform_string), decreasing = TRUE)

#plot of landforms
dotchart2(lf, col ='black', xlim = c(0, max(lf, cex.labels = 0.75)))
detach(package:Hmisc)

```


### Elevation, Slope and Aspect

```{r, echo=FALSE, results='asis', warning=FALSE}
vars <- c("elev_field", "slope_field")
morf <- subset(s, select = vars)
morf.lo <- melt(morf, measure.vars = vars)
morf.5n <- ddply(morf.lo, .(variable), summarize,
                range = prettySummary(value, p=c(0,0.5,1))
                )

if (!all(is.na(s$aspect_field))) {
  aspect <- subset(s, select = c("aspect_field"))
  aspect.lo <- melt(aspect, measure.vars = "aspect_field")
  aspect.lo$value <- circular(aspect.lo$value, template="geographic", units="degrees", modulo="2pi")
  aspect.5n <- ddply(aspect.lo, .(variable), summarize,
                range = prettySummary(value, p=c(0,0.5,1))
                )
  kable(rbind(morf.5n, aspect.5n), caption = "Elevation, slope gradient and aspect (min, median, max)(number of pedons)", align = "c")
  } else(kable(morf.5n, caption="Elevation and slope gradient (min, median, max)(number of pedons)", align = "c"))         

#bwplot(~ value | variable, data = morf.lo, 
#       main = "Boxplots of elevation and slope gradient",
#       scales=list(x="free"), axis = axis.grid,
#       as.table = TRUE
#       )
```

### Parent Material vs. Landform

```{r parent material}
if (!all(is.na(s[c("pmorigin", "pmkind", "landform_string")]))) {
  pm_comb <- factor(paste0(s$pmorigin, " ", s$pmkind))
  pm_land <- factor(s$landform_string)
  pm.lf <- addmargins(table(pm_comb, pm_land))
  kable(pm.lf, caption="Parent material vs landform (counts)")
  } else "The parent material and landform fields are empty."
```

### Slope Shape

```{r shape}
if (!all(is.na(s[c("shapedown", "shapeacross")]))) {
  kable(addmargins(xtabs(~ shapedown + shapeacross, data = s)), caption = "Down slope (y-axis) vs across slope (x-axis) (counts)")
  } else "The slope shape fields are empty."
```

### Hillslope Position vs. Drainage Class

```{r hillslope position}
if (any(complete.cases(s[c("hillslopeprof", "drainagecl")]))) {
  kable(addmargins(xtabs(~ drainagecl + hillslopeprof, data = s)), digits = 0, caption = "Drainage class vs hillslope position (counts)")
  } else "The hillslope position and drainage class fields are empty."
```